---
title: "HDAT9800 Chapter 5 Optional Unmarked Exercise"
author: "<insert your name here>"
date: "xx/xx/2024"
output:
  html_document:
    self_contained: false
    includes:
      in_header: header.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(leaflet)
```

## Introduction

This is an exercise in embedding _leaflet_ maps in a markdown document.

Complete the following tasks. The exercise is umarked and does not attarct any marks for the HDAT9800 course.

You will need to load any libraries you need in the setup chunk.
If your solution needs a library that has not been used so far this semester
(it should not need to) please include dependencies and instructions and be aware
that if it can't be made to run, the marks available are severely diminished.
Play it safe and don't be too fancy!

## Task 1: A basic ma

Include a basic map below that has a marker at the Sydney GPO labelled 'Sydney GPO'.
The Sydney GPO is at 33°52′04″S 151°12′28″E.

```{r gpo}
# include your GPO map here
```

## Task 2: Custom icons

Code (from the `header.html` file) has been included in the header of this tutorial
(look at the `output:` section of the YAML meta-data at the top of the document)
to include a font which has icons in it.

A value for the `icons` argument to _addAwesomeMarkers()_ can be constructed using

```
icons <- awesomeIcons(icon="envelope", library="fa")
```

replace the _addMarkers()_ line in your pipeline with _addAwesomeMarkers()._
Use the code above to define an `icons` variable
and use it as an argument to `_addAwesomeMarkers()` to add the icon.
The other arguments to _addAwesomeMarkers()_ are the same as for _addMarkers()._

Add extra `iconColor` and `markerColor` arguments to _awesomeIcons()_ to make
the icon colour `#fff` and the marker colour `red`.

```{r white-gpo}
# include your custom marker map here
```

## Loading some spatial data

We're going to load some spatial data with polygons for 2011 postcode boundaries from the ABS.
So you don't have to load extra libraries we're already loaded the shape files provided
by the ABS and saved the resulting data frames in two files:

  * `ABS_postcodes_2011.rds`
  * `ABS_postcodes_2011@data.rds`

To load these you we run the following code in a chunk.
The shape information is actually split into two different data frames which must both be loaded.

```{r load_polygons}
postcodes <- readRDS("ABS_postcodes_2011.rds")
postcodes@data <- readRDS("ABS_postcodes_2011@data.rds")
```

Once you have done this you can look at the data (without the polygon information) as follows

```{r examine_data}
head(postcodes@data)
```

We're interested in the `POA_NAME` field.

To extract the polygon for a particular postcode we can run:

```
# the postcode for the suburbs Paddington and Centennial Park is 2021
paddington <- postcodes[postcodes$POA_NAME=="2021",]
```

## Task 3: Highlighting polygons

  * extract the polygons for Kensington (postcode 2033) and UNSW (postcode 2052)
  * add a map with a polygon for Kensington shaded in the default colour and for UNSW in `red`

`addPolygons(map, data=polygon)` will add the polygon data in `polygon`.

```{r kensington}
# include your polygon map here
```

## Task 4: Interactivity 

Add the information in tasks 2 and 3 into one map with an iteractive menu
using _addLayersControl()._
The menu should allow the following three layers to be turned on or off:

  * the GPO marker
  * the Kensington boundary
  * the UNSW boundary
  
You will need to add a `group` argument to the three layers in order to control them.
  
Add the options `layersControlOptions(collapsed = FALSE)` to the layer controls.

```{r interactive}
# include your interactive map here
```
