---
title: "assignment_3"
author: "Zhenyu_Zhang(z5037788)"
format: html
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

### Introduction
The dataset used in this assignment is a retrospective heart failure dataset using electronic health data collected from patients who were admitted to Zigong Fourth People’s Hospital in Sichuan, China between December 2016 and June 2019. The dataset includes 168 variables for 2,008 patients with heart failure.
Data on subsequent hospital admissions and mortality were obtained at a mandatory follow‐up visit at 28 days, 3 months and 6 months (if the patient was unable to reach the clinical centre, the follow‐up visit was replaced by a telephone call).

Heart failure was defined according to the European Society of Cardiology (ESC) criteria:
- The presence of symptoms and/or signs of heart failure. Typical symptoms include breathlessness, orthopnea, paroxysmal nocturnal dyspnea, reduced excises  tolerance, Fatigue, tiredness, increased time to recover after exercise and Ankle swelling. Typical signs include elevated jugular venous pressure, hepatojugular reflux, third heart sound (gallop rhythm) and laterally displaced apical impulse.
- Elevated levels of BNPs (BNP >35 pg/mL and/or NT‐proBNP >125 pg/mL)
- Objective evidence of other cardiac functional and structural alterations underlying heart failure.
- In case of uncertainty, a stress test or invasively measured elevated LV filling pressure may be needed to confirm the diagnosis.


```{r libraries, echo=FALSE}
library(readr)
library(janitor)
library(ggplot2)
library(tidyverse)
library(knitr)
library(dplyr)
library(ggridges)
library(tidyr)
library(patchwork)
library(shiny)
```


```{r dataset, echo=FALSE}
# Load the data
dataset <- readr::read_csv("../zigong/dat.csv")

# Clean column names
dataset_cn <- dataset %>% janitor::clean_names()

# Convert relevant columns to factors
dataset_cf <- dataset_cn %>%
  mutate(
    gender = as.factor(gender),
    destination_discharge = as.factor(destination_discharge),
    admission_ward = as.factor(admission_ward),
    admission_way = as.factor(admission_way),
    occupation = as.factor(occupation),
    discharge_department = as.factor(discharge_department),
    type_of_heart_failure = as.factor(type_of_heart_failure),
    nyha_cardiac_function_classification = as.factor(nyha_cardiac_function_classification),
    killip_grade = as.factor(killip_grade),
    type_ii_respiratory_failure = as.factor(type_ii_respiratory_failure),
    consciousness = as.factor(consciousness),
    respiratory_support = as.factor(respiratory_support),
    oxygen_inhalation = as.factor(oxygen_inhalation),
    outcome_during_hospitalization = as.factor(outcome_during_hospitalization),
    age_cat = factor(age_cat, labels = c("21-29", "29-39", "39-49", "49-59", "59-69", "69-79", "79-89", "89-110"))
  )

# Filter dataset to remove unrealistic height and weight values
dataset_cleaned <- dataset_cf %>%
  filter(weight >= 30 & weight < 200, height >= 1.2 & height < 2.2)
```

---

```{r shiny_1, echo=FALSE}
shinyApp(
  ui1 <- fluidPage(
    titlePanel("Distribution of Heart Failure Types by Gender"),
    sidebarLayout(
      sidebarPanel(
        selectInput("gender", "Select Gender:",
                    choices = c("All", "Female", "Male"), selected = "All"),
        numericInput("y_min", "Y-axis Minimum:", value = 0, min = 0, max = 1000, step = 1),
        numericInput("y_max", "Y-axis Maximum:", value = 1000, min = 1, max = 2000, step = 1),
        radioButtons("y_scale", "Y-axis Scale:",
                     choices = c("Normal" = "normal", "Logarithmic" = "log"), 
                     selected = "normal")
      ),
      mainPanel(
        plotOutput("barPlot")
      )
    )
  ),
  
  server1 <- function(input, output) {
    output$barPlot <- renderPlot({
      data <- dataset_cleaned
      if (input$gender != "All") {
        data <- data[data$gender == input$gender, ]
      }
      
      p <- ggplot(data, aes(x = type_of_heart_failure, fill = gender)) +
        geom_bar(position = "dodge") +
        labs(title = "Distribution of Heart Failure Types by Gender", 
             x = "Type of Heart Failure", 
             y = "Count", 
             fill = "Gender") +
        scale_fill_manual(values = c("Female" = "pink", "Male" = "lightblue")) +
        theme_minimal()
      
      if (input$y_scale == "log") {
        # Ensure y_min and y_max are valid for log scale
        y_min_log <- ifelse(input$y_min <= 0, 1, input$y_min)
        y_max_log <- ifelse(input$y_max <= 0, 100, input$y_max)
        
        p <- p + scale_y_log10() + 
          annotation_logticks(sides = "l") +
          coord_cartesian(ylim = c(y_min_log, y_max_log))
      } else {
        p <- p + coord_cartesian(ylim = c(input$y_min, input$y_max))
      }
      
      p
    })
  }
)

```
---

```{r shiny_2, echo=FALSE}
shinyApp(
  ui2 <- fluidPage(
    titlePanel("LVEF by Type of Heart Failure and Gender"),
    sidebarLayout(
      sidebarPanel(
        selectInput("gender", "Select Gender:",
                    choices = c("All", "Female", "Male"), selected = "All"),
        checkboxGroupInput("age_group2", "Select Age Groups",
                           choices = c("21-29", "29-39", "39-49", "49-59", "59-69", "69-79", "79-89", "89-110"),
                           selected = c("39-49", "49-59", "59-69", "69-79"))
      ),
      mainPanel(
        plotOutput("scatterPlot")
      )
    )
  ),
  
  server2 <- function(input, output) {
    output$scatterPlot <- renderPlot({
      data <- dataset_cleaned
      if (input$gender != "All") {
        data <- data[data$gender == input$gender, ]
      }
      if (!is.null(input$age_group2)) {
        data <- data[data$age_cat %in% input$age_group2, ]
      }
      
      ggplot(data, aes(x = type_of_heart_failure, y = lvef, color = gender)) +
        geom_jitter(width = 0.2, height = 0) +
        geom_hline(yintercept = 40, linetype = "dashed", color = "#B76776") +
        geom_hline(yintercept = 50, linetype = "dashed", color = "#B76776") +
        labs(title = "LVEF by Type of Heart Failure and Gender", 
             x = "Type of Heart Failure", 
             y = "LVEF (%)", 
             color = "Gender") +
        annotate("text", x = 3.4, y = 33, label = "HFrEF", color = "#403E4E", vjust = -1) +
        annotate("text", x = 3.4, y = 43, label = "HFmrEF", color = "#403E4E", vjust = -1) +
        annotate("text", x = 3.4, y = 53, label = "HFpEF", color = "#403E4E", vjust = -1) +
        theme_minimal()
    })
  }
)


```

---

```{r shiny_3, echo=FALSE}
shinyApp(
  ui3 <- fluidPage(
    titlePanel("Blood Pressure Distribution by Gender within Selected Age Range"),
    sidebarLayout(
      sidebarPanel(
        selectInput("gender", "Select Gender:",
                    choices = c("All", "Female", "Male"), selected = "All"),
        sliderInput("age_range", "Select Age Range",
                    min = 21, max = 110, value = c(30, 80), step = 1)
      ),
      mainPanel(
        plotOutput("densityPlot")
      )
    )
  ),
  
  server3 <- function(input, output) {
    output$densityPlot <- renderPlot({
      data <- dataset_cleaned
      if (input$gender != "All") {
        data <- data[data$gender == input$gender, ]
      }
      data <- data %>%
        filter(between(as.numeric(sub(".*-", "", age_cat)), input$age_range[1], input$age_range[2]))
      
      long_dataset_bp <- data %>%
        pivot_longer(cols = c(systolic_blood_pressure, diastolic_blood_pressure, map), 
                     names_to = "blood_pressure_type", values_to = "value")
      
      ggplot(long_dataset_bp, aes(x = value, color = blood_pressure_type, linetype = gender)) +
        geom_density(aes(group = interaction(blood_pressure_type, gender)), adjust = 1.5) +
        labs(title = "Blood Pressure Distribution by Gender within Selected Age Range", 
             x = "Blood Pressure (mmHg)", 
             y = "Density", 
             color = "Blood Pressure Type", 
             linetype = "Gender") +
        scale_color_manual(values = c("systolic_blood_pressure" = "#B76776", 
                                      "diastolic_blood_pressure" = "#2F76C0", 
                                      "map" = "#7F807F"),
                           labels = c("systolic_blood_pressure" = "Systolic Blood Pressure", 
                                      "diastolic_blood_pressure" = "Diastolic Blood Pressure", 
                                      "map" = "Mean Arterial Pressure")) +
        scale_linetype_manual(values = c("Female" = "solid", "Male" = "dashed")) +
        theme_minimal()
    })
  }
)

```








