---
title: "HDAT9800 Chapter 3 Optional Unmarked Exercise"
author: "Your name"
date: "XX/XX/2024"
output: html_document
---

```{r setup, include=FALSE}
# include any libraries you need here, you will at least need the following
library(datasets)
library(ggplot2)
# set any global chunk options using knitr::opts_chunk$set(...)
```

## Introduction

We will be looking at the `ChickWeight` dataset from the _datasets_ package.

The `ChickWeight` dataset contains `r nrow(ChickWeight)` observations of `r length(names(ChickWeight))` variables.

You can read more about `ChickWeight` by typing `?ChickWeight` in the console when _datasets_ is loaded.

Let's have a look at the first few lines of `ChickWeight`.

```{r head}
head(ChickWeight)
```

The description of `ChickWeight` is as follows:

> Weight versus age of chicks on different diets

Let's plot some aspects of `ChickWeight` to have a look.

## Discussion

First we'll plot `Time` _versus_ `weight` for the entire dataset.

```{r time-versus-weight}
ggplot(ChickWeight, aes(x=Time, y=weight)) +
  geom_point()
```

We'd expect `weight` to go up over time.
Let's add a smoothed mean to the graph using `geom_smooth()`. 

```{r time-versus-weigh-with-mean}

```

Perhaps the `Diet` might affect `weight` gain.
Let's differentiate our data points by changing the colour according to `Diet`. 

```{r time-versus-weigh-with-mean-by-diet}

```

That's a bit hard to interpret.
Facet the graphs by `Diet` as well so that each different value of `Diet` has its own plot. 

```{r time-versus-weigh-with-mean-facetted-by-diet}

```

Add a layer to the graph which indicates mean ± standard error for each diet.
Draw these in black so they can be more easily seen. 

```{r time-versus-weigh-with-mean-facetted-by-diet-with-mean-and-se}

```

Let's tidy this up a bit by adding units to the axis labels.
Add `(g)` to the y-axis label and `(days)` to the x-axis label. 

```{r time-versus-weigh-with-mean-facetted-by-diet-labelled}

```

Finally, add a title to the plot: `Chick weight over time by diet`. 

```{r time-versus-weigh-with-mean-facetted-by-diet-labelled-with-title}

```

Out of curiosity, let's try facetting by `Chick` instead (but leaving the colour set by `Diet`) :

```{r time-versus-weigh-with-mean-facetted-by-chick-labelled, warning=FALSE}

```

## Questions

  1. From the graphs, which diet seems to result in improved weight gain?
  2. From the graphs, which chick(s) may have died during the course of the experiment? 
