---
title: "HDAT9600 Final Team Assignment"
subtitle: "Please see course timetable / 'Announcements' for submission deadline"
author: "insert your team name here"
date: "insert date of completion here"
output: html_document
---

```{r setup, include=FALSE}
# leave this code here, but feel free to adjust the options or add some more
# see the knitr documentation for details
knitr::opts_chunk$set(echo = TRUE, fig.width=12, fig.height=12)
```

## Instructions

* This file (hdat9600assess4_teamX.Rmd) is the R Markdown document in which you need to complete your HDAT9600 final team assignment.
* You should rename the file by replacing 'X' with your team number.
* This assignment is assessed and will count for 30% of the total course marks.
* The assignment comprises two tasks. The first task will focus on building a model to PREDICT an outcome and the second task will focus on using a model to EXPLAIN the relationships between a variable of interest and the outcome.
* There is no word limit, nor limit on the length of your submitted Rmarkdown (i.e. this file) document. However, the **rendered** html document, were it to be printed, should be **no more** than about 10 A4 pages in length. 
* Your html rendered document should include sufficient code chunks, output, and visual display as necessary to support your discursive commentary addressing each task. You may therefore wish to **suppress some of the contents** of your Rmarkdown file from appearing within the rendered document to ensure your output focuses on the most relevant results and discussion to address the questions.
* The rendered (html) document will be the submission primarily considered when awarding credit. However, the Rmarkdown file may also be considered if required to understand the analysis.
* <strong> Excessively long reports will be penalised </strong> by cessation of awarding of credit after roughly 10 pages of content.
* Feel free to delete the assignment instructions as you see fit to allow you some more room. 
* All tasks are intended to be completed as a team. Have fun discussing the tasks and working together.

Don't hesitate to ask the course convenor for help via OpenLearning. The course instructor is happy to point you in the right direction and to make suggestions, but they won't, of course, complete your assessment for you!


## Data for this assessment

The data used for this assessment consist of records from Intensive Care Unit (ICU) hospital stays in the USA. All patients were adults who were admitted for a wide variety of reasons. ICU stays of less than 48 hours have been excluded. 

The source data for the assessment are data made freely available for the 2012 MIT PhysioNet/Computing for Cardiology Challenge. Details are provided [here](https://physionet.org/challenge/2012/). Training Set A data have been used. The original data has been modified and assembled to suit the purpose of this assessment. 

The dataframe consists of 120 variables, which are defined as follows:

#### Patient Descriptor Variables
<li> <em>RecordID:</em>            a unique integer for each ICU stay</li>
<li> <em>Age:</em>                 years</li>
<li> <em>Gender:</em>              male/female</li>
<li> <em>Height:</em>              cm</li>
<li> <em>ICUType:</em>             Coronary Care Unit; Cardiac Surgery Recovery Unit; Medical ICU; Surgical ICU</li>
<li> <em>Length_of_stay:</em>      The number of days between the patient’s admission to the ICU and the end of hospitalisation</li>
<li> <em>Survival:</em>            The number of days between ICU admission and death for patients who died</li>


#### Outcome Variables
<li> <em>in_hospital_death:</em>   0:survivor/1:died in-hospital **this is the outcome variable for Task 1: Logistic Regression**</li>
<li> <em>Status:</em>              True/False **this is the censoring variable for Task 2: Survival Analysis**</li>
<li> <em>Days:</em>                Length of survival (in days) **this is the survival time variable for Task 2: Survival Analysis**</li>


#### Clinical Variables
<em>Use the hyperlinks below to find out more about the clinical meaning of each variable.</em>
The first two clinical variables are summary scores that are used to assess patient condition and risk.

* SAPS-I score - Simplified Acute Physiological Score [Le Gall et al., 1984](http://www.ncbi.nlm.nih.gov/pubmed/6499483)
* SOFA score - Sequential Organ Failure Assessment [Ferreira et al., 2001](http://www.ncbi.nlm.nih.gov/pubmed/11594901)

###

The following 36 clinical measures were assessed at multiple timepoints during each patient's ICU stay. For each of the 36 clinical measures, you are given 3 summary variables: a) The minimum value during the first 24 hours in ICU (_min), b) The maximum value during the first 24 hours in ICU (_max), and c) The difference between the mean and the most extreme values during the first 24 hours in ICU (_diff). For example, for the clinical measure Cholesterol, these three variables are labelled 'Cholesterol_min', 'Cholesterol_max', and 'Cholesterol_diff'.

* [Albumin](http://en.wikipedia.org/wiki/Human_serum_albumin)
 (g/dL)
* [ALP](http://en.wikipedia.org/wiki/Alkaline_phosphatase)
 [Alkaline phosphatase (IU/L)]
* [ALT](http://en.wikipedia.org/wiki/Alanine_transaminase)
 [Alanine transaminase (IU/L)]
* [AST](http://en.wikipedia.org/wiki/Aspartate_transaminase)
 [Aspartate transaminase (IU/L)]
* [Bilirubin](http://en.wikipedia.org/wiki/Bilirubin)
 (mg/dL)
* [BUN](http://en.wikipedia.org/wiki/BUN)
 [Blood urea nitrogen (mg/dL)]
* [Cholesterol](http://en.wikipedia.org/wiki/Cholesterol)
 (mg/dL)
* [Creatinine](http://en.wikipedia.org/wiki/Serum_creatinine#Plasma_creatinine)
 [Serum creatinine (mg/dL)]
* [DiasABP](http://en.wikipedia.org/wiki/Diastolic_blood_pressure)
 [Invasive diastolic arterial blood pressure (mmHg)]
* [FiO2](http://en.wikipedia.org/wiki/FIO2)
 [Fractional inspired O<sub>2</sub> (0-1)]
* [GCS](http://en.wikipedia.org/wiki/Glasgow_coma_score)
 [Glasgow Coma Score (3-15)]
* [Glucose](http://en.wikipedia.org/wiki/Serum_glucose)
 [Serum glucose (mg/dL)]
* [HCO3](http://en.wikipedia.org/wiki/Bicarbonate#Diagnostics)
 [Serum bicarbonate (mmol/L)]
* [HCT](http://en.wikipedia.org/wiki/Hematocrit)
 [Hematocrit (%)]
* [HR](http://en.wikipedia.org/wiki/Heart_rate)
 [Heart rate (bpm)]
* [K](http://en.wikipedia.org/wiki/Hypokalemia)
 [Serum potassium (mEq/L)]
* [Lactate](http://en.wikipedia.org/wiki/Lactic_acid)
 (mmol/L)
* [Mg](http://en.wikipedia.org/wiki/Magnesium#Biological_role)
 [Serum magnesium (mmol/L)]
* [MAP](http://en.wikipedia.org/wiki/Mean_arterial_pressure)
 [Invasive mean arterial blood pressure (mmHg)]
* [MechVent](http://en.wikipedia.org/wiki/Mechanical_ventilation)
 [Mechanical ventilation respiration (0:false, or 1:true)]
* [Na](http://en.wikipedia.org/wiki/Serum_sodium)
 [Serum sodium (mEq/L)]
* [NIDiasABP](http://en.wikipedia.org/wiki/Diastolic_blood_pressure)
 [Non-invasive diastolic arterial blood pressure (mmHg)]
* [NIMAP](http://en.wikipedia.org/wiki/Mean_arterial_pressure)
 [Non-invasive mean arterial blood pressure (mmHg)]
* [NISysABP](http://en.wikipedia.org/wiki/Systolic_blood_pressure)
 [Non-invasive systolic arterial blood pressure (mmHg)]
* [PaCO2](http://en.wikipedia.org/wiki/Arterial_blood_gas)
 [partial pressure of arterial CO<sub>2</sub> (mmHg)]
* [PaO2](http://en.wikipedia.org/wiki/Arterial_blood_gas)
 [Partial pressure of arterial O<sub>2</sub> (mmHg)]
* [pH](http://en.wikipedia.org/wiki/Arterial_blood_gas)
 [Arterial pH (0-14)]
* [Platelets](http://en.wikipedia.org/wiki/Platelets)
 (cells/nL)
* [RespRate](http://en.wikipedia.org/wiki/Respiratory_physiology)
 [Respiration rate (bpm)]
* [SaO2](http://en.wikipedia.org/wiki/Arterial_blood_gas)
 [O<sub>2</sub> saturation in hemoglobin (%)]
* [SysABP](http://en.wikipedia.org/wiki/Systolic_blood_pressure)
 [Invasive systolic arterial blood pressure (mmHg)]
* [Temp](http://en.wikipedia.org/wiki/Normal_human_body_temperature)
 [Temperature (&deg;C)]
* [TropI](http://en.wikipedia.org/wiki/Troponin)
 [Troponin-I (&mu;g/L)]
* [TropT](http://en.wikipedia.org/wiki/Troponin)
 [Troponin-T (&mu;g/L)]
* [Urine](http://en.wikipedia.org/wiki/Fluid_balance)
 [Urine output (mL)]
* [WBC](http://en.wikipedia.org/wiki/Reference_ranges_for_blood_tests#Hematology)
 [White blood cell count (cells/nL)]
* Weight (kg)


## Accessing the Data
There are two datasets provided. The data frames can be loaded with the following code:

```{r}
#update with your own filepath
#mydat <- file.path("C:/Users/z3520358/OneDrive - UNSW/0HDAT9600_T1_2024/Group Project")

#icu_patients_df0 <- readRDS(file.path(mydat,"icu_patients_df0.rds"))
#icu_patients_df1 <- readRDS(file.path(mydat,"icu_patients_df1.rds"))

```

**Note:** icu_patients_df1 is an imputed (i.e. missing values are 'derived') version of icu_patients_df0.  This assessment does not concern the methods used for imputation. You should use `icu_patients_df1` for all tasks unless it is specifically noted to use `icu_patients_df0`. 


## Aims of your project

You have two analytic goals in this assignment. 
1. The first goal (Task 1) is to build a model to PREDICT in-hospital death. The context for this is that the hospital management wishes to use this model to prioritise the allocation of resources with the overarching aim of improving patient survival.  
2. The second goal (Task 2) is to try to understand and EXPLAIN how survival varies by `Age`. How does survival change with increasing Age? Are age-related survival differences explained solely by differing clinical characteristics of patients? 


# Task 1

### Part A: Exploratory data analysis

Conduct a brief EDA for the dataset `icu_patients_df1`. One of your goals for this EDA should be to identify a sub-set of variables (approx 15-20 variables) that _may_ be useful predictors of survival and/or relevant to your analytical task. You may wish to firstly undertake a brief literature review to try to identify factors that may be important predictors of survival in an ICU. This doesn't need to be in-depth, but enough to get a basic idea of what the variables are measuring and which variables might, in theory, be most important. 

For the 15-20 chosen variables, present enough information to: (i) explain why you selected these variables, based on a combination of your literature review and EDA, and (ii) Provide an overview of the variables and how they are related to `in-hospital death`. Present your summary using tables and/or plots with supporting commentary where relevant. 

### Part B: Predictive logistic model

Your aim is to build a model to PREDICT in-hospital death. In this task, you are required to develop a logistic regression model using the `icu_patients_df1` data set which adequately **predicts** the `in_hospital_death` variable as the outcome and utilises predictor variables from your chosen subset. You should fit a series of models, evaluating each one, before you present your final model. Your final model should **not** include all the predictor variables, just a small selection of them, which you have selected based on statistical significance and/or background knowledge. Remember, your aim is prediction, but you should also consider generalisability of the model, so we are aiming for parsimony. Aim for between five and ten predictor variables (slightly more or fewer is OK). You should assess each model you consider for goodness of fit and other relevant statistics to help you choose between them. For your final model, present a set of relevant diagnostic statistics and/or charts and comment on them. 

Finally, re-fit your final model to the unimputed data frame (`icu_patients_df0.rds`) and comment on any differences you find compared to the same model fitted to the imputed data. 


**Create your response to task 1 here, as a mixture of embedded (`knitr`) R code and any resulting outputs, and explanatory or commentary text. Add code chunks as you see fit and choose whether you wish for the code and or results to be displayed in the final html document.** 

```{r}



```




# Task 2

### Part A: Exploratory Data Analysis

The goal of this EDA is similar to Task 1 and you may choose to use the same sub-set of predictors if they are relevant and appropriate to the analysis for Task 2. You may also choose to select a different sub-set if you wish. Justify whichever approach you choose.

Present enough information to: (i) explain why you selected these variables, and (ii) Provide an overview of the variables and how they are related to survival. Present your summary using tables and/or plots with supporting commentary where relevant.


### Part B: Explanatory survival model

Your aim is to try to understand and EXPLAIN how survival time varies by `Age`. How does survival change with increasing Age?  Can any age-related differences in survival time be explained by differing clinical presentations of older patients?

In this task, you are required to develop a Cox proportional hazards survival model using the `icu_patients_df1` data set which adequately **explains** the length of survival indicated by the `Days` variable, with censoring as indicated by the `Status` variable.  You should fit a univariable model examining the relationship between Age and survival. You should then also fit another one or two multivariable models, to examine the extent that survival is explained by other variables in the dataset and whether age is independently related to survival after controlling for other factors.  Your final model should **not** include all the predictor variables, just a relevant subset of them, which you have selected based on statistical significance and/or background knowledge.  It is perfectly acceptable to include predictor variables in your final model which are not statistically significant, as long as you justify their inclusion on medical or physiological grounds (you will not be marked down if your medical justification is not exactly correct, but do you best). You should assess each model you consider for goodness of fit and other relevant statistics, and you should assess your final model for violations of assumptions and perform other diagnostics which you think are relevant (and modify the model if indicated, or at least comment on the possible impact of what your diagnostics show). 

Finally, re-fit your final model to the unimputed data frame (`icu_patients_df0.rds`) and comment on any differences you find.

**Create your response to task 2 here, as a mixture of embedded (`knitr`) R code and any resulting outputs, and explanatory or commentary text. Add code chunks as you see fit and choose whether you wish for the code and or results to be displayed in the final html document.** 

```{r}

```




### Hints

1. Make sure you justify your choice of potential predictors. This could be based on a range of different factors, eg. review of the literature, data quality assessment, strong association with the outcome. 

2. Present the story of your analysis by including explanation and commentary supported by well-presented plots and/or tables. 

3. Make sure all plots and tables are clearly presented with titles and labels as needed. Pay attention to sizing of plots and make sure you check how the rendered result appears. 

4. Clearly define your starting population for the analysis and describe the characteristics of your outcome variable. 

5. Present summaries of appropriate univariate models examining the unadjusted relationship between each predictor and the outcome. 

6. Fit an appropriate series of multivariable regression models, justifying your approach. Assess each model you consider for goodness of fit and other relevant statistics.

7. Present your final model for each task and justify why this model is the most appropriate to answer the analytical question(s). Your final model(s) should **not** include all the predictor variables, just a small subset of them, which you have selected based on statistical significance, background knowledge and relevance to achieving the analytical aim.

8. For your _final model_, present a set of diagnostic statistics and/or charts and comment on them.

9. For each task make sure you write a paragraph or two summarising the most important findings of your analysis. Include reference to the most important results from the statistical output, and a simple clinical interpretation. Make sure your summary of findings addresses the main analytical aim(s) for each task. 




## Save, knit and submit

**Reminder**: don't forget to save this file, to knit it to check that everything works and appears in the right format, and then submit TWO documents via the drop box in OpenLearning (.rmd AND .html).


### Problems?

If you encounter problems with any part of the process described above, please contact the course convenor via OpenLearning as soon as possible so that the issues can be resolved in good time, and well before the assessment is due.


### Additional Information

The instructions are deliberately less prescriptive than the individual assessments to allow you some latitude in what you do and how you go about the task. However, to complete the tasks, you only need to replicate or repeat the steps covered in the course - you do not need to attempt any types of analyses that have not been covered. Refer to the marking rubric for how marks will be awarded.   

Note also that with respect to the model fitting, there are no **right** or **wrong** answers when it comes to variable selection and other aspects of model specification. Deep understanding of the underlying medical concepts which govern patient treatment and outcomes in ICUs is not required or assumed, although you should try to gain some understanding of each variable using the links provided. You will not be marked down if your medical justifications are not exactly correct or complete, but do you best, and don't hesitate to seek help from the course convenor.   
